---
title: "Reproducible Research, Assignment 2"
author: "Aditya Salapaka"
output: html_document
---

## About This Assignment
Storms and other severe weather events can cause both public health and economic
problems for communities and municipalities. Many severe events can result in 
fatalities, injuries, and property damage, and preventing such outcomes to the 
extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) storm database. This database tracks characteristics of 
major storms and weather events in the United States, including when and where 
they occur, as well as estimates of any fatalities, injuries, and property 
damage.

## About The Data

[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

National Weather Service [Storm Data Documentation] (https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In 
the earlier years of the database there are generally fewer events recorded, 
most likely due to a lack of good records. More recent years should be 
considered more complete.

## Title


## Data Acquisition
The data was downloaded from the link provided above.

```{r cache = TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
              "stormdata.csv.bz2")

data <- read.csv(bzfile("stormdata.csv.bz2")) # cache = TRUE
```

Since the questions that needed to be answered required only damage to property
and crops, a subset of the relevant data was taken, and the original data-frame
was removed from the memory.

```{r}
subsetData <- subset(data, select = c("BGN_DATE","EVTYPE","PROPDMG",
                                            "PROPDMGEXP","CROPDMG","CROPDMGEXP",
                                            "FATALITIES","INJURIES"))
rm(data)
```

## Clean up
Data before 1996 was incomplete, so data before this time was removed. Only 
records past 1-1-1996 were considered.

Initially, the date in the data frame had to be cleaned up. For this, the time
was removed as it seemed unchanged, and therefore, irrelevant. 

```{r}
BGN_DATE <- subsetData$BGN_DATE
BGN_DATE <- substr(BGN_DATE, 1, 9)
subsetData$BGN_DATE <- as.Date(BGN_DATE, format = "%m/%d/%Y")

subsetData <- subset(subsetData, subsetData$BGN_DATE > as.Date("1995-12-31"))

```

The `PROPDMGEXP` and `CROPDMGEXP` columns contained exponents for `PROPDMG` 
and `CROPDMG`. These exponents were converted to their respective numbers and 
multiplied with the data to arrive at final values to work with. These were
stored in columns `PROPDMGEXPTOT` and `CROPDMGEXPTOT`.

For this purpose, a function `exponent` was created.

```{r}
exponent <- function (x){
        x <- as.character(x)
        ifelse(x == "B", as.numeric(1000000000), 
               ifelse(x == "M", as.numeric(1000000), 
                      ifelse(x == "K", as.numeric(1000), 0)))
}
```

subsetData$PROPDMGEXP <- toupper(subsetData$PROPDMGEXP)
subsetData$PROPDMGEXP <- exponent(subsetData$PROPDMGEXP)
subsetData$PROPDMGEXPTOT <- as.numeric(subsetData$PROPDMG*subsetData$PROPDMGEXP)

subsetData$CROPDMGEXP <- toupper(subsetData$CROPDMGEXP)
subsetData$CROPDMGEXP <- exponent(subsetData$CROPDMGEXP)
subsetData$CROPDMGEXPTOT <- as.numeric(subsetData$CROPDMG*subsetData$CROPDMGEXP)

The data documentation has defined 48 types of events which are:

```
Astronomical Low Tide
Avalanche
Blizzard
Coastal Flood
Cold/Windhill
Debris Flow
Dense Fog
Dense Smoke
Drought
Dust Devil
Dust Storm
Excessive Heat
Extremeold/Windhill
Flash Flood
Flood
Frost/Freeze
Funnelloud
Freezing Fog
Hail
Heat
Heavy Rain
Heavy Snow
High Surf
High Wind
Hurricane (Typhoon)
Ice Storm
Lake-Effect Snow
Lakeshore Flood
Lightning
Marine Hail
Marine High Wind
Marine Strong Wind
Marine Thunderstorm Wind
Ripurrent
Seiche
Sleet
Storm Surge/Tide
Strong Wind
Thunderstorm Wind
Tornado
Tropical Depression
Tropical Storm
Tsunami
Volcanic Ash
Waterspout
Wildfire
Winter Storm
Winter Weather
```

The number of unique events were then found out.

```{r}
subsetData$EVTYPE <- toupper(subsetData$EVTYPE)

unique <- length(unique(subsetData$EVTYPE))
```

This gives `r unique` number of events. This has to be reduced to 48.

First, that part of the data was taken care of  where the `EVTYPE` is similar to
the 48 events listed. To check how many events are left, a function was 
created `similarevents`. It accepts FALSE or TRUE to give events which either 
match the 48 events, or are different from them.

```{r}
similarevents <- function(x){
        unique(grep("ASTRONOMICAL LOW TIDE|AVALANCHE|BLIZZARD|COASTAL FLOOD|
        COLD/WINDHILL|DEBRIS FLOW|DENSE FOG|DENSE SMOKE|DROUGHT|DUST DEVIL|
        DUST STORM|EXCESSIVE HEAT|EXTREMEOLD/WINDHILL|FLASH FLOOD|FLOOD|
        FROST/FREEZE|FUNNELLOUD|FREEZING FOG|HAIL|HEAT|HEAVY RAIN|HEAVY SNOW|
        HIGH SURF|HIGH WIND|HURRICANE (TYPHOON)|ICE STORM|LAKE-EFFECT SNOW|
        LAKESHORE FLOOD|LIGHTNING|MARINE HAIL|MARINE HIGH WIND|
        MARINE STRONG WIND|MARINE THUNDERSTORM WIND|RIPURRENT|SEICHE|SLEET|
        STORM SURGE/TIDE|STRONG WIND|THUNDERSTORM WIND|TORNADO|
        TROPICAL DEPRESSION|TROPICAL STORM|TSUNAMI|VOLCANIC ASH|WATERSPOUT|
        WILDFIRE|WINTER STORM|WINTER WEATHER", subsetData$EVTYPE, 
                    value = TRUE, invert = x))
        }
```

A text file with all 48 events was created, which is `events.txt`.

The events which were similar, but not same, as the 48 events were then found
out. 

```{r}
events <- readLines("events.txt")
events <- toupper(events)

print(sort(str_trim(setdiff(similarevents(FALSE), events))))
```

These were then replaced by the event names they correspond to.

```{r}
for (i in 1:length(x)){
        subsetData$EVTYPE[grepl(x[i], subsetData$EVTYPE)] <- x[i]
}
```

All these events have been reduced to their standard types.
```{r}
print(sort(str_trim(setdiff(similarevents(FALSE), events))))
```

```{r}
unique <- length(unique(subsetData$EVTYPE))
```

There are now `r unique` number of events remaining. We have to tackle the other
events which are not similar to the 48 events.

The other events were found out by `r similarevents(FALSE)`. Data cleanup was
started in alphabetical order, starting from "ASTRONOMICAL LOW TIDE" and ending
with "WINTER WEATHER".

A function was created, `findevents`, which would find events with a particular
term. For example, `findevents("TIDE")` would give all events with "tide" in
them.

```{r}
findevent <- function(x){
        unique(grep(x, subsetData$EVTYPE, value = TRUE))
}
```
